TacticToe: Learning the Game of Tactics

### 2018-01-17 tactictoe_jv.tex
tactictoe: case study examples

\usepackage[final]{listings}
%\usepackage[scaled=.95]{couriers}
\lstdefinelanguage{SML}%
{columns=fullflexible,
  keywords={THEN,val,open,store_thm,fun,fn},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\footnotesize\selectfont,%
  basicstyle=\fontfamily{pcr}\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {=>}{{$\Rightarrow$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {!}{$\forall$}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }

\lstdefinelanguage{SMLSmall}%
{columns=fullflexible,
  keywords={THEN,val,open,store_thm,fun,fn},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\scriptsize\selectfont,%
  basicstyle=\fontfamily{pcr}\small\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {=>}{{$\Rightarrow$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {!}{$\forall$}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }
\lstset{language=SML}
\begin{document}
\maketitle

@@

\section{Introduction}\todo{TG + JU + CK}
%\begin{example} Proof automatically generated by \tactictoe for the user given 
%goal
%\small
%\begin{center}
%
%
%\begin{BVerbatim}[commandchars=\\\{\}]
%Goal: ``\(\forall\)l. FOLDL (\(\lambda\)xs x. SNOC x xs) [] l = l``
%Proof:
%  SNOC_INDUCT_TAC THENL
%  [ REWRITE_TAC [APPEND_NIL, FOLDL],
%    ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC]
%      THEN CONV_TAC (DEPTH_CONV BETA_CONV)
%      THEN ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC] ]
%\end{BVerbatim}
%
%
%\end{center}
%
%\end{example}
%
\vspace{100mm}
\tableofcontents

@@

\begin{table}[b!]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} 
& {meas} \\
\midrule
\tactictoe & 74.6 & 67.3 & 77.3 & 28.2\\
\eprover & 50.8 & 67.6 & 63.4 & 10.8 \\
\midrule
\phantom{abc}  & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 37.3 & 71.4 & 58.4 & 73.6 \\
\eprover & 19.3 & 20.7 & 15.8 & 20.9 \\
\bottomrule
\end{tabular}
\caption{\label{theories}Percentage (\%) of re-proved theorems in the theories 
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},  
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}. 
}
\end{table}

@@

(Todo: change next figure to an histogram)
\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=size, y=solved] {data/tt_percent};
\addplot table[x=size, y=solved] {data/e_percent};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original 
proof}
\end{figure}

@@

\section{Case study}\todo{TG: rerun experiment with prettification and 
minimization + add another example. 
  TG + RK: describe the two examples + find examples that did not appear during 
  evaluation
  RK: After implementation and small experiment is done (in about 1 week) 
      Test tactictoe on new goals and manual exploration mode.
      Give feedback.
}
%Investigating further the different qualities of \tactictoe, we study its 
%generated proof scripts on an example in \texttt{list} theory (see 
%Example~\ref{ex:casestudy}). 
%The theorem to be proven states the equivalence 
%between the fact that a number $n$ is greater than the length of a list 
%$\mathit{ls}$ with the 
%fact that dropping $n$ elements from this list returns an empty list.
%
%The human proof proceeds by induction on $n$ followed by solving both goals 
%using rewrite steps combined with an arithmetic decision procedure. 
%Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
%induction but solve 
%the base case and the inductive case in a different way.
%The base case only needs rewriting using the global simpset in the 
%\tactictoe(NH)
%proof, which is simulated by a call to \metis in the (SH) proof.
%The inductive case should require an arithmetic decision procedure as hinted 
%by 
%the human 
%proof. This is achieved by rewriting using an arithmetic simpset in the second 
%proof. In the first proof however, a rewriting step and case splitting step 
%were used
%to arrive at a point where \metis calls succeed.
%The tactic proof produced by \tactictoe(NH) often looks better than the one 
%discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
%large numbers of premises.

@@

@@

%SURJ_INJ_INV
%proof length: 7
%proof time: 0.092014
%Statistics
%  infstep : 500
%  nodes   : 132
%  maxdepth: 7
%Time: 9.574603
%  inferstep time: 7.287698
%  node_find time: 0.135171
%  node_crea time: 1.674807
%    pred time: 0.335816
%    thmpred time: 0.000018
%    mc time: 0.704148
%    inst time: 0.0

@@

\begin{example}\label{ex:casestudy}(
Proof discovered by \tactictoe for theorem \texttt{SURJ\_INJ\_INV} In 
theory \texttt{pred\_set})
\small
\begin{center}
\begin{alltt}
    SRW_TAC [] [auto] 
  THEN 
    POP_ASSUM (MP_TAC o 
      CONV_RULE (DEPTH_CONV RIGHT_IMP_EXISTS_CONV 
                 THENC SKOLEM_CONV THENC REWRITE_CONV [EXISTS_DEF] 
                 THENC DEPTH_CONV BETA_CONV)) 
  THEN
    Q.SPEC_TAC (`@y. !x. x IN t ==> y x IN s \wed (f (y x) = x)`,`g`) 
  THEN
    SRW_TAC [] [INJ_DEF] 
  THEN
    Q.EXISTS_TAC `g` 
  THEN
    SRW_TAC [] [INJ_DEF] 
  THEN
    METIS_TAC [auto]

\end{alltt} 
\end{center}
\end{example}

>>>>

\usepackage[final]{listings}
%\usepackage[scaled=.95]{couriers}
\lstdefinelanguage{SML}%
{columns=fullflexible,
  keywords={THEN,THENL,val,open,store_thm,fun,fn},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\footnotesize\selectfont,%
  basicstyle=\fontfamily{pcr}\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {==>}{{$\Rightarrow$}}1%
   {wedge}{{$\wedge$}}1%
   {>=}{{$\geq$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {!}{$\forall$}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }

\lstdefinelanguage{SMLSmall}%
{columns=fullflexible,
  keywords={THEN,THENL,val,open,store_thm,fun,fn},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\scriptsize\selectfont,%
  basicstyle=\fontfamily{pcr}\small\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {==>}{{$\Rightarrow$}}1%
   {wedge}{{$\wedge$}}1%
   {<=}{{$\leq$}}1%
   {>=}{{$\geq$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {!}{$\forall$}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }
\lstset{language=SML}
\begin{document}
\maketitle

@@

\section{Introduction}\todo{TG + JU + CK}
%\begin{example} Proof automatically generated by \tactictoe for the user given 
%goal
%\small
%\begin{center}
%
%
%\begin{BVerbatim}[commandchars=\\\{\}]
%Goal: ``\(\forall\)l. FOLDL (\(\lambda\)xs x. SNOC x xs) [] l = l``
%Proof:
%  SNOC_INDUCT_TAC THENL
%  [ REWRITE_TAC [APPEND_NIL, FOLDL],
%    ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC]
%      THEN CONV_TAC (DEPTH_CONV BETA_CONV)
%      THEN ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC] ]
%\end{BVerbatim}
%
%
%\end{center}
%
%\end{example}
%
\vspace{100mm}\tableofcontents

@@

\begin{table}[]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} 
& {meas} \\
\midrule
\tactictoe & 74.6 & 67.3 & 77.3 & 28.2\\
\eprover & 50.8 & 67.6 & 63.4 & 10.8 \\
\midrule
\phantom{abc}  & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 37.3 & 71.4 & 58.4 & 73.6 \\
\eprover & 19.3 & 20.7 & 15.8 & 20.9 \\
\bottomrule
\end{tabular}
\caption{\label{theories}Percentage (\%) of re-proved theorems in the theories 
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},  
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}. 
}
\end{table}

@@

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=size, y=solved] {data/tt_percent};
\addplot table[x=size, y=solved] {data/e_percent};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original 
proof}
\end{figure}

@@

\section{Case study}\todo{
  TG + RK: describe the two examples.
  RK: After implementation and small experiment is done (in about 1 week) 
      Test tactictoe find new examples on new goals and manual exploration mode.
      Give feedback.
}
%Investigating further the different qualities of \tactictoe, we study its 
%generated proof scripts on an example in \texttt{list} theory (see 
%Example~\ref{ex:casestudy}). 
%The theorem to be proven states the equivalence 
%between the fact that a number $n$ is greater than the length of a list 
%$\mathit{ls}$ with the 
%fact that dropping $n$ elements from this list returns an empty list.
%
%The human proof proceeds by induction on $n$ followed by solving both goals 
%using rewrite steps combined with an arithmetic decision procedure. 
%Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
%induction but solve 
%the base case and the inductive case in a different way.
%The base case only needs rewriting using the global simpset in the 
%\tactictoe(NH)
%proof, which is simulated by a call to \metis in the (SH) proof.
%The inductive case should require an arithmetic decision procedure as hinted 
%by 
%the human 
%proof. This is achieved by rewriting using an arithmetic simpset in the second 
%proof. In the first proof however, a rewriting step and case splitting step 
%were used
%to arrive at a point where \metis calls succeed.
%The tactic proof produced by \tactictoe(NH) often looks better than the one 
%discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
%large numbers of premises.

@@

%SURJ_INJ_INV
%proof length: 2
%proof time: 0.050853
%Statistics
%  infstep : 31
%  nodes   : 11
%  maxdepth: 3
%Time: 1.300789
%  inferstep time: 1.071869
%  node_find time: 0.018046
%  node_crea time: 0.147604
%    pred time: 0.04263
%    thmpred time: 0.000005
%    mc time: 0.071282
%    inst time: 0.0
%Proof found: 
%Org tac number: 15
%Replaying proof: 0.562829

@@

@@

Proofs discovered by \tactictoe discovered during evaluation.

for theorem \texttt{SURJ\_INJ\_INV} In 
theory \texttt{pred\_set})

\begin{example}\label{ex:casestudy}()

%val SURJ_INJ_INV = store_thm(
%  "SURJ_INJ_INV",
%  ``SURJ f s t ==> ?g. INJ g t s /\ !y. y IN t ==> (f (g y) = y)``,
\vspace{5mm}
\small

Goal
\begin{lstlisting}[language=SMLSmall,frame=tb]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
REWRITE_TAC [IMAGE_SURJ] THEN
  DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
  BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC
  THENL [
  irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
    REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
    REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
    FULL_SIMP_TAC std_ss [THE_DEF],
  ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
    RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
      [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC [] ]
\end{lstlisting}

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}

Goal
\begin{lstlisting}[language=SMLSmall,frame=tb]
!l1 l2 n x. LENGTH l1 <= n ==> 
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}

%val LUPDATE_APPEND2 = Q.store_thm("LUPDATE_APPEND2",
%   `!l1 l2 n x.
%      LENGTH l1 <= n ==>
%      (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n-LENGTH l1) l2))`,
%);
\vspace{5mm}

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}

If a discovered proof is reasonable short, fast and understandable, they could 
replace its original in the \holfour standard library.

%
%LUPDATE_APPEND2
%proof length: 6
%proof time: 0.017156
%Statistics
%  infstep : 70
%  nodes   : 27
%  maxdepth: 4
%Time: 3.02716
%  inferstep time: 2.12842
%  node_find time: 0.07146
%  node_crea time: 0.508384
%    pred time: 0.164047
%    thmpred time: 0.000012
%    mc time: 0.265314
%    inst time: 0.0
%Proof found: 
%Org tac number: 10
%Original proof time: 0.0
%Replaying proof: 0.470118

\end{example}
